---
title: "Writing More Efficient Code"
editor: source
execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(tictoc)
library(palmerpenguins)

options(digits = 2)
```

In this module, you'll learn a bit about how to make your code more efficient 
**and** more human readable. 

# Code Efficiency

For you, most of the work to make your code more efficient (and more readable)
can be largely summarized with two principles:

1. avoid larger calculations and memory demands, and 
2. avoid repeating slow steps

Let's explore how we can write code which addresses each of these common 
pitfalls. 

# Avoid larger calculations and memory demands

In many of our analyses, we will find ourselves performing repeated tasks on 
the same, smaller data frame. For example, consider the following dataset:

```{r}
#| echo: true
#| message: false
#| label: load-data

library(fivethirtyeight)

glimpse(congress_age)
```

Suppose we want to do some exploratory analysis on these data, summarizing 
and visualizing the ages of senators. Let's use the `tic()` and `toc()`
functions from the **tictoc** packages to check how much time it takes to run
this code:

```{r}
#| results: false
#| message: false
#| fig-show: hide
#| echo: true
#| label: exploratory-analysis

tic()

congress_age %>%
  filter(chamber == "senate") %>%
  summarize(mean(age))

congress_age %>%
  filter(chamber == "senate") %>%
  slice_max(n = 5, order_by = age)

congress_age %>%
  filter(chamber == "senate") %>%
  ggplot() + 
  geom_histogram(mapping = aes(x = age))
```

```{r}
#| echo: true
toc()
```

These are all three reasonable things to do, and they can't be done in the same
pipeline. But wait, we just made R do the process of filtering to only include
senators three separate times!

```{r}
#| echo: true
#| results: false

tic()
  congress_age %>%
    filter(chamber == "senate")
```

```{r}
#| echo: true
toc()
```

Instead, how about we filter the data once, save it into an object, and call on
that object:

```{r}
#| echo: true
#| fig-show: hide
#| message: false
#| results: false

tic()

 senate <- congress_age %>%
  filter(chamber == "senate")

senate %>%
  summarize(mean(age))

senate %>%
  slice_max(n = 5, order_by = age)

senate %>%
  ggplot() + 
  geom_histogram(mapping = aes(x = age))
```

```{r}
#| echo: true

toc()
```

Not only did saving the senator data cut down on the runtime of our code, it 
made the code easier to understand! It is **much** clearer to the reader that
each of the data exploration steps refers to the `senate` dataset. 

::: {.callout-check-in}
1. What functions help us assess how long it takes our code to run?

2. How would you change the code below to make it more efficient?

```{r}
#| echo: true
#| eval: false
#| label: repeated-steps-quiz

penguins |> 
  group_by(species) |>
  summarize(mean_bill_len = mean(bill_length_mm, na.rm = TRUE))

penguins |> 
  group_by(species) |>
  slice_min(order_by = bill_length_mm)

penguins |> 
  group_by(species) |> 
  mutate(
    scaled_bill_length = (bill_length_mm - mean(bill_length_mm, na.rm = TRUE)) /
      sd(bill_length_mm, na.rm = TRUE)
    )

```
:::

# Avoid repeating slow steps

## Avoid for-loops

`for()`-loops are deathly slow in R. If you absolutely must iterate over a
process, rely on the `map()` or `apply()` function families. These functions 
offer the same versatility of a `for()`-loop but are (1) more human readable, 
and (2) much faster because they are written in C. 

::: {.callout-note}
# Why choose `map()` instead of `apply()`? 

The `map()` and `apply()` functions will typically be close to the same speed, 
so the largest  argument in favor of the `map()` family of functions are
their consistent syntax, and their consistent object output. 

The `map()` function **always** returns a list and the `map_XXX()` variants
**always** return a vector of a specified data type. Whereas, `apply()` returns
a vector or a list, `lapply()` returns a list, and `sapply()` returns a vector
or a matrix. 
:::

Here is a motivating example demonstrating the difference in run time between 
two functions:

1. a function which loops over every column in a data frame to extract the data
type of that column
2. a function which uses `map_chr()` to apply the `class()` function to
every column of a data frame

```{r}
#| echo: true
#| label: map-apply-functions

loop_func <- function(df){
  
  # initialize the type object
  type <- rep(NA, ncol(df))
  
  for(i in seq_len(ncol(df))) {
    
    # store the class of the ith column of df into the ith entry of type 
    type[i] <- class(df[,i])
  }
  
  return(type)
}

map_func <- function(df){
  
  type <- map_chr(df, class)
  
  return(type)

  }
```

Okay, now that we've made these two functions, let's see how they compare in 
their run time. Below, we've made a simple albeit large dataframe. The dataframe
has five rows and 100,000 columns and the element of every column is a vector
of 1s. 

```{r}
#| label: simple-large-dataframe
#| echo: true

dat <- as.data.frame(
  matrix(1,
         nrow = 5,
         ncol = 100000)
  )

```

Let's use `tic()` and `toc()` again to compare the run time for these 
two functions:

::: columns
::: {.column width="48%"}
```{r}
#| warning: false
#| echo: true
#| label: loop-function-speed-1
#| results: false

tic()
loop_func(dat)
```

```{r}
#| label: loop-function-speed-2
#| echo: true

toc()
```
:::

::: {.column width="4%"}
:::

::: {.column width="48%"}
```{r}
#| echo: true
#| results: false
#| label: map-function-speed-1

tic()
map_func(dat)
```

```{r}
#| echo: true
#| label: map-function-speed-2

toc()
```
:::
:::

Looking at the output, we see substantial differences in the run time of these
two functions. The loop function took about 3.5 seconds whereas the map function
took less than half a second. 

::: {.callout-check-in}
3. Why might we prefer using `map()` functions instead of writing a `for()` 
loop?

a. `map()` automatically makes the code run faster in all cases.
b. `map()` makes code more readable, concise, and often faster.
c. `map()` can only be used with numeric data, unlike loops.
d. `map()` replaces all possible uses of loops in programming.

4. Which `map()` function should be used in place of this `for()` loop?

```{r}
#| eval: false
#| echo: true
#| label: loop-quiz

numbers <- c(4, 9, 16, 25)

results <- vector("numeric", length(numbers))

for (i in seq_along(numbers)) {
  results[i] <- sqrt(numbers[i])
}
```

a. `map()`
b. `map_dbl()`
c. `map_chr()`
d. `walk()`
:::

## Use vectorized functions

Better even than `map()` or `apply()` is not to iterate at all! Writing 
vectorized functions is tough, but do-able in many cases. Here's an example:

::: {.callout-required-video}
::: youtube-video-container
{{< video https://youtu.be/ShewFCbPHv8?si=sVpKCo0FF28hw40v >}}
:::
:::

::: {.callout-check-in}
5. Which of the following functions are vectorized?

a. `sqrt()`
b. `if()`
c. `ifelse()`
d. `mean()`

6. Why does the code below produce an error?

```{r}
#| echo: true
#| error: true

x <- c(1, 2, 3, 4, 16, 25)

f <- function(x) {
  if (x > 0) x else 0
}

f(x)
```
:::
