---
title: "Making Great Tables with gt"
editor: source
---

```{r}
#| label: packages
#| message: false
#| warning: false

library(tidyverse)
library(palmerpenguins)
library(gt)
```

## Formatting Tables

This week, we will practice making nice, report worthy, tables!

I would recommend you think of tables no different from the visualizations
you've been making. We want **all** aspects of our tables to be clear to the
reader, so the comparisons we want them to make are straightforward. You should
be thinking about:

- Column headers
- Grouping headers
- Order of columns
- Order of rows
- Number of decimals included for numeric entries
- etc.

Tables are
also a great avenue to display creativity! In fact, there is a yearly
[RStudio table contest](https://www.rstudio.com/blog/rstudio-table-contest-2022/), 
and [here](https://community.rstudio.com/c/table-gallery/64) is a gallery of the
award winning tables!

There are many packages for generating tables but we recommend the [**gt** 
(greatables) R package](https://gt.rstudio.com/). 

For simple tables, the basic `gt()` function from the **gt** package works 
quite well. For more sophisticated tables, you will need to add-on additional
styling functions (e.g., `cols_label()`, `tab_header()`, `fmt_percent()`).

<!-- ::: callout-warning -->
<!-- Quarto doesn't play nice with some options for formatting HTML tables in other -->
<!-- packages.  -->

<!-- To make sure that your tables render as expected, we need to specify  -->
<!-- `html-table-processing: none` in the YAML header.  -->
<!-- ::: -->

## Making Simple Tables

Let's explore the gt package using the `palmerpenguins` dataset. We can start
with a simple table with information about the dataset. Let's display the 
data type of each variable in the `palmerpenguins` dataset:

```{r}
#| label: first-gt

penguins |> 
  map_chr(.f = class) |> 
  enframe(name = "Variable", value = "Data Type") |> 
  gt()
```

You'll notice that by default, gt produces a nicely formatted HTML table. If I
wanted to make the table more professional, I would think about adding the 
following features:

- rows ordered to make the information easy to understand
- bolded column names
- a table caption or header

```{r}
#| label: gt-boldface-headers
penguins |> 
  map_chr(.f = class) |> 
  enframe(name = "Variable", value = "Data Type") |> 
  arrange(`Data Type`) |> 
  gt() |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
    ) |> 
  tab_header(
    title = md("Data Types of Variables in the `penguins` Dataset")
  ) |> 
  tab_caption(
    caption = md("Example 2: `gt` table with simple extras.")
    )
```

::: {.callout-tip}
# But wait there's more!

You'll notice in the [documentation for `tab_style()`](https://gt.rstudio.com/reference/tab_style.html) that there are
**many** additional styling options (e.g., color) that we're not using here. 
:::

## Make More Complex Tables

To demonstrate making a more complex table, we're going to translate 
[this article from the gt package](https://gt.rstudio.com/articles/case-study-gtcars.html)
into the `penguins` data.

### Row Groups

To display observations related to a specific group, we can use `group_by()`
before passing the data into `gt()`. This step creates a row for each unique
value of `species`:

```{r}
#| label: gt-row-groups
penguins |> 
  group_by(species) |>
  slice_head(n = 10) |> 
  gt()
```


Similar to what I did before, I could arrange this table based on a variable:

```{r}
#| label: gt-row-groups-arrange

penguins |> 
  group_by(species) |>
  arrange(body_mass_g) |> 
  slice_head(n = 10) |> 
  gt()

```

## Hiding Columns & Moving Columns

We can use our handy friend the `select()` function to remove columns we 
aren't interested in (`-c(sex, year, island)`) and to reorder the columns 
we want to keep. Now it is much easier to see that the table is arranged based 
on `body_mass_g` since it is the first variable that appears in the table!

```{r}
#| label: hide-move

penguins |> 
  select(-c(sex, year, island), 
         body_mass_g, 
         bill_depth_mm, 
         bill_length_mm, 
         flipper_length_mm) |> 
  group_by(species) |>
  arrange(body_mass_g) |> 
  slice_head(n = 10) |> 
  gt()
```

## Using Formatter Functions

The gt package also comes with powerful functions for reformatting the text
displayed in the columns. For example, if you are working with data on
currencies, the `fmt_currency()` function would help you format the column so 
every value has the necessary currency strings (e.g., `$100.35`). Similarly, 
the `fmt_percent()` will format every value in a column to have a `%` symbol
after the value (e.g., `75%`).

To explore these functions, we will need to do a bit of data summarizing. Let's
calculate the mean for each of the penguin measurement columns. 

::: {.column-margin}
::: {.callout-important}
# Efficiency & Readability

Notice that I'm calculating the mean for **four** different columns. It would 
be difficult to read, cumbersome to type, and error prone to use four 
different lines within `summarize()` to do these calculations. Instead, let's
use the `across()` function!
:::
:::

```{r}
#| label: tbl-decimals
penguins |> 
  select(-c(sex, year, island),
         species,
         body_mass_g, 
         bill_depth_mm, 
         bill_length_mm, 
         flipper_length_mm) |> 
  group_by(species) |>
  summarize(
    across(.cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g), 
           .fns = ~mean(.x, na.rm = TRUE), 
           .names = "{.col}_mean"
           )
    ) |> 
  gt()

```

The decimal values for these columns are a bit too long. Let's use this as an
opportunity to explore the `fmt_number()` function:

```{r}
#| label: tbl-formatted-decimals

penguins |> 
  select(-c(sex, year, island), 
         body_mass_g, 
         bill_depth_mm, 
         bill_length_mm, 
         flipper_length_mm) |> 
  group_by(species) |>
  summarize(
    across(.cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g), 
           .fns = ~mean(.x, na.rm = TRUE), 
           .names = "{.col}_mean"
           )
    ) |> 
  gt() |> 
  fmt_number(columns = -species, 
             decimals = 2)
```

Much better! Now let's see if we can clean up the column headers!

## Putting Columns Into Groups

All of the measurements in this table are related to various aspects of a 
penguins morphology: bill, flipper, body mass. It seems like this could be a 
great place to play around with adding groups to these columns. 

```{r}
#| label: tab-spanner

penguins |> 
  select(-c(sex, year, island), 
         body_mass_g, 
         bill_depth_mm, 
         bill_length_mm, 
         flipper_length_mm) |> 
  group_by(species) |>
  summarize(
    across(.cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g), 
           .fns = ~mean(.x, na.rm = TRUE), 
           .names = "{.col}_mean"
           )
    ) |> 
  gt() |> 
  fmt_number(columns = -species, 
             decimals = 2) |> 
  tab_spanner(
    label = "Bill Measurements",
    columns = c(bill_length_mm_mean, bill_depth_mm_mean)
  ) |> 
  tab_spanner(
    label = "Flipper Measurements",
    columns = flipper_length_mm_mean
  ) |> 
  tab_spanner(
    label = "Body Mass Measurements",
    columns = body_mass_g_mean
  ) 
  
```
Well, now that we have these groups, it seems like we could simplify the names
of each group's columns. 

```{r}
#| label: tab-spanner-better-column-names

penguins |> 
  select(-c(sex, year, island), 
         body_mass_g, 
         bill_depth_mm, 
         bill_length_mm, 
         flipper_length_mm) |> 
  group_by(species) |>
  summarize(
    across(.cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g), 
           .fns = ~mean(.x, na.rm = TRUE), 
           .names = "{.col}_mean"
           )
    ) |> 
  gt() |> 
  fmt_number(columns = -species, 
             decimals = 2) |> 
  tab_spanner(
    label = "Bill Measurements",
    columns = c(bill_length_mm_mean, bill_depth_mm_mean)
  ) |> 
  tab_spanner(
    label = "Flipper Measurements",
    columns = flipper_length_mm_mean
  ) |> 
  tab_spanner(
    label = "Body Mass Measurements",
    columns = body_mass_g_mean
  ) |> 
  cols_label(
    bill_length_mm_mean = "Mean Length (mm)",
    bill_depth_mm_mean = "Mean Depth (mm)",
    flipper_length_mm_mean = "Mean Length (mm)", 
    body_mass_g_mean = "Mean Mass (g)", 
    species = "Penguin Species"
    )
  
```

### Adding Color

For our last exploration, let's add some color to the table! The `data_color()`
function allows us to add colors to the cells of our table. The colors can be 
added to *every* cell (like shown below) or only added to *select* cells (based
on a condition that is checked for the rows). 

```{r}
#| label: table-with-color

penguins |> 
  select(-c(sex, year, island), 
         body_mass_g, 
         bill_depth_mm, 
         bill_length_mm, 
         flipper_length_mm) |> 
  group_by(species) |>
  summarize(
    across(.cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g), 
           .fns = ~mean(.x, na.rm = TRUE), 
           .names = "{.col}_mean"
           )
    ) |> 
  gt() |> 
  fmt_number(columns = -species, 
             decimals = 2) |> 
  tab_spanner(
    label = "Bill Measurements",
    columns = c(bill_length_mm_mean, bill_depth_mm_mean)
  ) |> 
  tab_spanner(
    label = "Flipper Measurements",
    columns = flipper_length_mm_mean
  ) |> 
  tab_spanner(
    label = "Body Mass Measurements",
    columns = body_mass_g_mean
  ) |> 
  cols_label(
    bill_length_mm_mean = "Mean Length (mm)",
    bill_depth_mm_mean = "Mean Depth (mm)",
    flipper_length_mm_mean = "Mean Length (mm)", 
    body_mass_g_mean = "Mean Mass (g)", 
    species = "Penguin Species"
    ) |> 
  data_color(
    method = "numeric",
    palette = "PuOr", 
    reverse = TRUE
  )
```

## Now go forth and make great tables!

::: {.callout-check-in}
1. The `tab_style()` function allows you to modify which aspects of a table?

- title
- subtitle
- text of the cells
- fill of the cells
- borders of the cells
- column labels
- row group labels
- footnotes
- source notes
- none of the above
- all of the above

2. To specify multiple styles in the `style` argument of `tab_style()`, the 
styles must be specified as a [vector / dataframe / list]. 

3. Is it possible to specify multiple `locations` for the `styles` to be
applied?

- Yes
- No
- Only if they can be selected with the same `cells_XXXX()` function. 

4. There are two ways to create row groups in `gt()`. One option is to use 
`group_by()` [before / after] inputting the data into `gt()`. The second option
is to use `tab_row_group()` [before / after] inputting the data into `gt()`.

5. The `cols_label()` function is the only way to change the names of columns
in a `gt()` table. 
:::
