---
title: "Iterating Over Functions"
editor: source
---

In this unit, you’ll review for iteration---repeatedly performing the same
function on different inputs. 

Iteration in R generally tends to look a bit different from other programming
languages. Much of iteration we get for free! For example, if you want to double
a numeric vector `x` in R, you can just write `2 * x`, whereas in many other
languages you would need to explicitly double each element of `x` using some
sort of `for` loop.

In R, there are generally two methods for iteration---`for()` loops and 
functionals. We will start with a review of `for()` loops before hopping over
to functionals. 

```{r}
#| label: packages
#| include: false
library(tidyverse)
```

# `for` loops

In R, `for` loops have the following general structure:

```{r for_syntax, eval=FALSE}
for (i in some_vector) {
    # Code to do stuff with i
}
```

`some_vector` can be **any** vector, including:

- An indexing vector: `1:3`
- A character vector: `c("group1", "group2", "group3")`
- A vector of any other class

```{r}
#| label: for-examples

groups <- c("group1", "group2", "group3")

for (i in 1:3) {
    print(groups[i])
}

for (g in groups) {
    print(g)
}
```

## `for()` loop Indices

The `seq_along()` function generates an integer sequence from 1 to the length of
the vector supplied. A nice feature of `seq_along()` is that it generates an
empty iteration vector if the vector you're iterating over itself has length 0.

```{r}
#| label: for-seq-along

seq_along(groups)
no_groups <- c()
seq_along(no_groups)

for (i in seq_along(groups)) {
    print(groups[i])
}

for (i in seq_along(no_groups)) {
    print(no_groups[i])
}
```

Closely related to `seq_along()` is `seq_len()`. While `seq_along(x)` generates
an integer sequence from 1 to `length(x)`, `seq_len(x)` takes `x` itself to be a
length:

```{r}
#| label: for-seq-len

seq_len(3)
seq_len(0)

for (i in seq_len(length(groups))) {
    print(groups[i])
}

for (i in seq_len(length(no_groups))) {
    print(no_groups[i])
}
```

## Dataframe Indices

`seq_len()` is useful for iterating over the rows of a data frame because
`seq_along()` would iterate over columns:

```{r}
#| label: small-data

small_data <- tibble(a = 1:2, 
                     b = 2:3, 
                     c = 4:5)
small_data
```

```{r}
#| label: seq-along-length

for (col in seq_along(small_data)) {
    print(col)
}

for (r in seq_len(nrow(small_data))) {
    print(r)
}

```

## Storing Objects

Often we'll want to store output created during a `for` loop. We can create
storage containers with the `vector()` function:

```{r}
#| label: initialize-vectors

char_storage <- vector("character", 
                       length = 3)
char_storage

num_storage <- vector("numeric", 
                      length = 3)
num_storage

list_storage <- vector("list", 
                       length = 3)
list_storage
```

Our loop can then store objects created at each iteration stage:

```{r}
#| label: store-results

for (i in seq_len(3)) {
    char_storage[i] <- str_c("Number: ", i)
    num_storage[i] <- 2*i
    list_storage[[i]] <- i # Note the [[ for subsetting here
}

char_storage
num_storage
list_storage
```

::: {.callout-tip}
# Initial Object Size

Notice that we initialized `char_storage`, `num_storage`, and `list_storage` to
all be the same length as our iteration vector (`seq_len(3)`). Technically, this
wasn't necessary. We could have been much lazier when we initialized these 
objects: `list_storage <- NA`. Why didn't we do this?

When you don't initialize an object as a specific size (e.g., `length = 3`) 
then at every stage of the `for()` loop R needs to append a new index to that
object. This **really** slows down your `for()` loop!

**You should always initialize your object as the size you want it to be.**
:::

::: {.callout-important}
# Check-in: Iteration

1. Write a `for()`-loop that prints the even numbers from 1:20. 

2. Can you produce the same output with the `seq()` function?

```{r}
#| label: for-ex-solutions-1
#| echo: false
#| eval: false

for (i in seq_len(10)) {
    print(2*i)
}

seq(from = 2, to = 20, by = 2)
```


3. Write a `for()`-loop that iterates over the `month.name` vector (built-in to
base R) and stores a character vector of output containing strings like 
"Month 1: January", "Month 2: February". 

4. Can you produce the same output with `str_c()` only?

```{r}
#| label: for-ex-solutions-2
#| eval: false
#| echo: false

month_strings <- vector("character", length = length(month.name))

for (i in seq_along(month.name)) {
    month_strings[i] <- str_c("Month ", i, ": ", month.name[i])
}
month_strings

str_c("Month ", 1:12, ": ", month.name)
```

5. Write a `for()`-loop that store the `class()` (type) of every column in the
`mtcars` data frame. 

```{r}
#| label: col_classes_soln
#| eval: false
#| echo: false

col_classes <- vector("character", 
                      ncol(mtcars)
                      )

for (i in seq_along(mtcars)) {
    col_classes[i] <- class(mtcars[[i]])
}
```
:::

# Iteration with Functionals

> A functional is a function that takes a function as an input and returns a
> vector as output. 
> - Hadley Wickham

## 📖 Recommended Reading: [Functionals](https://adv-r.hadley.nz/functionals.html)

`purrr` is a `tidyverse` package that provides several useful functions for
iteration. The main advantages of `purrr` include:

- Improved readability of R code
- Reduction in the "overhead" in writing a `for()`-loop (creating storage
containers and writing the `for (i in ...)`)

In `purrr`, we can use the family of `map()` functions to apply a function to
each element of a list or vector. We can think of this as mapping an input (a
list or vector) to a new output via a function. Let's look at the [`purrr` cheatsheet](https://rstudio.github.io/cheatsheets/purrr.pdf) to look at
graphical representations of how these functions work.

- `map()` returns a list
- `map_chr()` returns a character vector
- `map_lgl()` returns a logical vector
- `map_int()` returns an integer vector
- `map_dbl()` returns a numeric vector
- `map_vec()` returns a vector of a different (non-atomic) type (like dates or
factors)

## A Single Output

To get the `class()` of each data frame column, `map_chr()` is the function we
want because the class of a variable is a strings (e.g., `"logical"`).

```{r}
#| label: purrr_col_classes

map_chr(mtcars, .f = class)
```

::: {.callout-note}
# List Input

The first input of `map()` functions must be a list. A dataframe is a special
type of list, where the columns are the different elements of the list 
(e.g., `mtcars[["mpg]]`). `map_chr()` iterates over the **columns** (elements) 
of the `mtcars` dataframe (list). 
:::

Let's get the class of each variable in `diamonds`:

```{r}
#| label:  purrr_col_classes_error
#| error: true

map_chr(diamonds, .f = class)
```

## Multiple Outputs

What happened!? `map_chr()` was expecting to create a character vector with 
**one** element per column in `diamonds`. But something happened in column 2
with the `cut` variable. Let's figure out what happened:

```{r}
#| label: class_diamonds_cut

class(diamonds$cut)
```

Ah! `cut` has **two** classes. In this case, `map()` (which returns a list) is
the best option because some variables have multiple classes:

```{r}
#| label: purrr_col_classes_no_error

map(diamonds, .f = class)
```

The error we encountered with `map_chr()` is a nice feature of `purrr` because
it requires us to be very sure of the type of output we are getting. Failing
loudly is vastly preferable to getting unexpected outputs silently!

## Combining with Tidy Selection

We can combine `map_*()` functions with tidy selection for some powerful
variable summaries that require much less code than `for()` loops.

```{r}
#| label:  purrr_examples_by_col

diamonds %>% 
  select(where(is.numeric)) %>% 
  map_dbl(.f = mean)

diamonds %>% 
  select(!where(is.numeric)) %>% 
  map_int(.f = n_distinct)
```

::: {.callout-important}
# Check-in: Iteration

6. Write a function called `summ_stats()` that takes a numeric vector `x` as
input and returns the mean, median, standard deviation, and IQR as a data frame.
You can use `tibble()` to create the data frame.
    * Example: `tibble(a = 1:2, b = 2:3)` creates a data frame with variables
    `a` and `b`.


```{r}
#| label: ex_summ_stats_soln1
#| echo: false

summ_stats <- function(x) {
    tibble(
        mean = mean(x, na.rm = TRUE),
        median = median(x, na.rm = TRUE),
        sd = sd(x, na.rm = TRUE),
        iqr = IQR(x, na.rm = TRUE)
    ) 
}
```

7. Use `map()` to apply your `summ_stats()` function to the numeric columns in 
the `diamonds` dataset. *Hint:* Look up the `bind_rows()` documentation from
dplyr to combine summary statistics for all quantitative variables into one
data frame. The `.id` argument will be especially helpful in adding the variable
names!

```{r}
#| label: ex_summ_stats_soln2
#| echo: false
#| eval: false

diamonds %>% 
  select(where(is.numeric)) |> 
  map(.f = summ_stats) %>% 
  bind_rows(.id = "variable") 

```

8. Let's make the output more intuitive, with the variables on the columns and
the summary statistics on the rows. You will need to do a double pivot 
(`pivot_longer()` then `pivot_wider()`) to achieve this result!

```{r}
#| label: double-pivot
#| echo: false
#| eval: false

diamonds %>% 
  select(where(is.numeric)) %>% 
  bind_rows(.id = "variable") %>% 
  pivot_longer(cols = -variable, 
               names_to = "statistic", 
               values_to = "value") %>% 
  pivot_wider(names_from = variable, values_from = value)

```
:::

## Multiple Inputs

`purrr` also offers the `pmap()` family of functions that take multiple inputs
and loops over them simultaneously. Let's look at the [`purrr` cheatsheet](https://rstudio.github.io/cheatsheets/purrr.pdf) to look at
graphical representations of how these functions work.

```{r}
#| label: pmap_ex1

string_data <- tibble(
    string = c("apple", "banana", "cherry"),
    pattern = c("p", "n", "h"),
    replacement = c("P", "N", "H")
)
string_data

pmap_chr(string_data, .f = str_replace_all)
```

Note how the column names in `string_data` exactly match the argument names in
`str_replace_all()`. The iteration that is happening is across rows, and the
multiple arguments in `str_replace_all()` are being matched by name.

We can also use `pmap()` to specify variations in some arguments but leave some
arguments constant across the iterations:

```{r}
#| label: pmap_ex2

string_data <- tibble(
    pattern = c("p", "n", "h"),
    replacement = c("P", "N", "H")
)

pmap_chr(string_data, str_replace_all, string = "ppp nnn hhh")
```

### Exercises

7. Create 2 small examples that show how `pmap()` works with `str_sub()`. Your
examples should:

- Use different arguments for `string`, `start`, and `end`
- Use different arguments for `start` and `end` but a fixed `string`

<details>
    <summary>Solution</summary>

```{r}
#| label: pmap_ex_str_remove_soln

string_data <- tibble(
    string = c("apple", "banana", "cherry"),
    start = c(1, 2, 4),
    end = c(2, 3, 5)
)

pmap_chr(string_data, str_sub)

string_data <- tibble(
    start = c(1, 2, 4),
    end = c(2, 3, 5)
)
pmap_chr(string_data, str_sub, string = "abcde")
```

</details>