---
title: "Web Scraping"
editor: source
execute: 
  warning: false
  message: false
---

<!-- **Part 4:** Obtain the following information for **all** cheeses in the -->
<!-- database: -->

<!-- -   cheese name -->
<!-- -   URL for the cheese's webpage (e.g., <https://www.cheese.com/gouda/>) -->
<!-- -   whether or not the cheese has a picture (e.g.,  -->
<!-- [gouda](https://www.cheese.com/gouda/) has a picture, but  -->
<!-- [bianco](https://www.cheese.com/bianco/) does not). -->

<!-- To be kind to the website owners, please add a 1 second pause between page -->
<!-- queries. (Note that you can view 100 cheeses at a time.) -->

<!-- **Part 5:** When you go to a particular cheese's page (like  -->
<!-- [gouda](https://www.cheese.com/gouda/)), you'll see more detailed information -->
<!-- about the cheese. For [**just 10**]{.underline} of the cheeses in the database, -->
<!-- obtain the following detailed information: -->

<!-- -   milk information -->
<!-- -   country of origin -->
<!-- -   family -->
<!-- -   type -->
<!-- -   flavour -->

<!-- (Just 10 to avoid overtaxing the website! Continue adding a 1 second pause -->
<!-- between page queries.) -->

<!-- **Part 6:** Evaluate the code that you wrote in terms of **efficiency**. To -->
<!-- what extent do your function(s) adhere to the **principles for writing good functions**? -->
<!-- To what extent are your **functions efficient**? To what extent is your  -->
<!-- **iteration of these functions efficient**?  -->

```{r}
#| label: setup
#| include: false

library(tidyverse)
```

# Learning goals

After this lesson, you should be able to:

-   Use CSS Selectors and the Selector Gadget tool to locate data of interest
within a webpage
-   Use the `html_elements()` and `html_text()` functions within the `rvest`
packages to scrape data from webpage using CSS selectors

# Web scraping

We have talked about how to acquire data from APIs. Whenever an API is available
for your project, you should default to getting data from the API. Sometimes an
API will not be available, and **web scraping** is another means of getting
data.

Web scraping describes the use of code to extract information displayed on a web
page. In R, the `rvest` package (meant to sound like "harvest") offers tools for
scraping.

## Scraping ethics

### `robots.txt`

`robots.txt` is a file that some websites will publish to clarify what can and
cannot be scraped and other constraints about scraping. When a website publishes
this file, this we need to comply with the information in it for moral and legal
reasons.

We will look through the information in [this tutorial](https://www.zenrows.com/blog/robots-txt-web-scraping) and apply this
to the [cheese.com robots.txt file](https://www.cheese.com/robots.txt).

The [cheese.com](https://www.cheese.com/) `robots.txt` is very simple, it
says that anyone is allowed to scrape (`User-agent: *`). The file doesn't give
any additional instructions, such as `Crawl-delay` (how long you need to wait
between scraping each page), `Visit-time` (restrictions on time of day that
scraping is allowed), or `Request-rate` (restrictions on simultaneous requests).

### Further considerations

The article [Ethics in Web Scraping](https://carpentry.library.ucsb.edu/2022-05-12-ucsb-webscraping/06-Ethics-Legality-Webscraping/index.html) 
describes some good principles to ensure that we are valuing the labor that 
website owners invested to provide data and creating good from the information
we do scrape.

# HTML structure

HTML (hypertext markup language) is the formatting language used to create
webpages. Let's look at the core parts of HTML from the [rvest vignette](https://cran.r-project.org/web/packages/rvest/vignettes/rvest.html).

## Finding CSS Selectors

In order to gather information from a webpage, we must learn the language used
to identify patterns of specific information. For example, on the 
[alphabetical page of cheeses]](https://www.cheese.com/alphabetical/), we can
see that the data is represented in a consistent pattern of image + name of 
cheese.

We will identify data in a web page using a pattern matching language called
[CSS Selectors](https://css-tricks.com/how-css-selectors-work/) that can refer
to specific patterns in HTML, the language used to write web pages.

For example:

-   Selecting by tag:
    -   `"a"` selects all hyperlinks in a webpage ("a" represents "anchor" links
    in HTML)
    -   `"p"` selects all paragraph elements
    
-   Selecting by ID and class:
    -   `".description"` selects all elements with `class` equal to
    "description"
        -   The `.` at the beginning is what signifies `class` selection.
        -   This is one of the most common CSS selectors for scraping because in
        HTML, the `class` attribute is extremely commonly used to format webpage
        elements. (Any number of HTML elements can have the same `class`, which
        is not true for the `id` attribute.)
    -   `"#mainTitle"` selects the SINGLE element with **id** equal to
    "mainTitle"
        -   The `#` at the beginning is what signifies `id` selection.

``` html
<p class="title">Title of resource 1</p>
<p class="description">Description of resource 1</p>

<p class="title">Title of resource 2</p>
<p class="description">Description of resource 2</p>
```

::: {.callout-warning}
Websites change often! So if you are going to scrape a lot of data, it is
probably worthwhile to save and date a copy of the website. Otherwise, you may 
return after some time and your scraping code will include all of the wrong CSS
selectors.
:::

Although you can [learn how to use CSS Selectors by hand](%5BCSS%20Selectors%5D(https://css-tricks.com/how-css-selectors-work/)), 
we will use a shortcut by installing the 
[Selector Gadget](http://selectorgadget.com/) tool.

-   There is a version available for Chrome--add it to Chrome via the [Chome Web Store](https://chromewebstore.google.com/detail/selectorgadget/mhjhnkcfbdhnjickkkdbjoemdmbfginb).
    -   Make sure to pin the extension to the menu bar. (Click the 3 dots \> 
    Extensions \> Manage extensions. Click the "Details" button under
    SelectorGadget and toggle the "Pin to toolbar" option.)
-   There is also a version that can be saved as a bookmark in the browser--see [here](https://rvest.tidyverse.org/articles/selectorgadget.html).

## Required Video: [Selector Gadget Tutorial Video](http://selectorgadget.com/)

## Check-in: Webscraping

Head over to the [alphabetical page of the first 100 cheeses](https://www.cheese.com/alphabetical/?per_page=100).
Click the Selector Gadget extension icon or bookmark button. As you mouse over
the webpage, different parts will be highlighted in orange.

1.  What CSS selector would you use to obtain the image of the cheese?

<!-- img -->

2.  What CSS selector would you use to obtain the name of the cheese?

<!-- h3 -->

3.  What CSS selector would you use to obtain the link for each cheese?

<!-- .product-item a -->

## Retrieving Data Using `rvest` and CSS Selectors

Now that we have identified CSS selectors for the information we need, let's
fetch the data using the `rvest` package.

```{r}
#| label: read_html
#| echo: true

library(rvest)

cheeses <- read_html("https://www.cheese.com/alphabetical/?per_page=100")
```

Once the webpage is loaded, we can retrieve data using the CSS selectors we
specified earlier. The code below retrieves three pieces of information about 
the cheeses on this website:

**1. name**

::: {.column-margin}
::: {.callout-tip}
# More Targetted CSS Selectors
I'm adding an `h3` to my CSS selector, since the store information is also
selected with the `.product-item`. If I don't grab the `h3` element, then 
the cheeses with store information will show up like this:
`"Stores >\n                \n            \n        \n    \n    Aged Gouda" `
:::
:::

```{r}
#| label: obtain-cheese-name

cheeses %>%
  html_elements(".product-item h3") %>%
  html_text(trim = T) |> 
  head()
```

**2. HTML**

```{r}
cheeses %>%
  html_elements(".product-item a") %>%
  html_attr("href") %>%
  url_absolute("https://www.cheese.com") |> 
  head()
```

::: {.column-margin}
::: {.callout-tip}
# Relative and Absolute URLs

The urls that are output by `html_attr("href")` are *relative* urls 
(e.g., `"/2-year-aged-cumin-gouda/"`) to get te *full* url, we need to paste the
base url (https://www.cheese.com) to the beginning of the relative url. The
`url_absolute()` function can do this for us. 

<!-- The function is essentially the same as: `str_c("https://www.cheese.com", .)`,  -->
<!-- where the string `"https://www.cheese.com"` is added to the beginning of every -->
<!-- relative url.  -->
:::
:::

**3. status of the image (exists or missing)**

```{r}
cheeses %>%
  html_elements(".product-item a img") %>%
  html_attr("class") %>%
  head()
```

::: {.column-margin}
::: {.callout-tip}
# Detecting an Image

This seems like a great place to use `str_detect()`!
:::
:::

```{r}
#| label: code from students
#| echo: false
#| eval: false

## VISRUTH & MATT
  # cheese product listings
  cheeses <- read_html(url) |>
    html_nodes(".product-item")

  # temp object that's used twice
  temp_cheese_data <- cheeses |>
    html_nodes("h3 a")

  # building tibble
  tibble(
    cheese = temp_cheese_data |> html_text(),
    url = temp_cheese_data |>
      html_attr("href") |>
      (\(slug) glue::glue("https://www.cheese.com{slug}"))(),
    has_image = cheeses |>
      html_nodes(".product-img a img") |>
      html_attr("class") |>
      str_detect("image-exists")
  )
```

## Iteration

Our goal is to obtain a dataset of characteristics about different cheeses 
from <https://www.cheese.com>. **Before doing any coding,** let's plan our
approach!

-   What functions will you write?
    -   We probably need a function to grab the text from the page.
    -   We can then use this function to grab different parts of an HTML page
    and piece them together
-   What arguments will they have?
    -   The function grabbing text should have two inputs, (1) the page and (2)
    the CSS selector to use
    -   The function scraping the page should probably only take one input: the
    page
-   How will you use your functions?
    -   These functions need to be iterated! We would need to call the function
    to scrape a page once for each page we want to scrape.

## Check-in: Webscraping

Let's carry out this plan to get the article title, publication date, and
abstract text for the first five pages of news releases in a single data frame.
You will need to write at least one function, and you will need
iteration---either a `for` loop or the appropriate `map_()` functions.

Advice:

-   Mouse over the page buttons at the very bottom of the news home page to see
what the URLs look like.
-   The abstract should not have the publication date--use `stringr` and regular
expressions to remove the publication date.
-   Include `Sys.sleep(2)` in your function to respect the `Crawl-delay: 2` in
the NIH `robots.txt` file.
-   Recall that `bind_rows()` from `dplyr` takes a list of data frames and
stacks them on top of each other.


```{r}
#| label: ex_solns_fcts
#| echo: true

get_text_from_page <- function(page, css_selector) {
    
  page %>%
    html_elements(css_selector) %>%
    html_text(trim = T)
}

get_url_from_page <- function(page, css_selector) {
    
  page %>%
    html_elements(css_selector) %>%
    html_attr("href")
}

get_image_from_page <- function(page, css_selector) {
    
  page %>%
    html_elements(css_selector) %>%
    html_attr("src")
}

scrape_page <- function(url) {
    
    # 1 second crawl delay
    Sys.sleep(1)
    
    # Read the page
    page <- read_html(url)
    
    # Grab elements from the page
    cheese_names <- get_text_from_page(page, ".product-item")
    cheese_url <- get_url_from_page(page, ".product-item a")
    cheese_image <- get_image_from_page(page, ".product-item img")
    
    # Clean cheese names
    cheese_names <- cheese_names %>%
      trimws()
    
    cheese_names <- ifelse(
      str_detect(cheese_names, "Stores >"),
      str_trim(str_extract(cheese_names, "[^\n]+$")),
      cheese_names
      )
    
    # Full cheese URL
    base_url <- "https://www.cheese.com"
    cheese_url <- paste0(base_url, cheese_url)
    cheese_url <- unique(cheese_url[!grepl("store", cheese_url)]) #remove store links + duplicates
    
    # Find cheeses with image
    has_image <- ifelse(grepl("static", cheese_image), "No", "Yes")
    
    #Put page elements into a dataframe
    cheeses <- data.frame(
      cheese = cheese_names,
      url = cheese_url,
      image = has_image
    )
    
    return(cheeses)
}
```

Using `purrr::map()`:

```{r}
#| eval: false
#| label:  ex_solns_map
#| echo: true

base_url <- "https://www.cheese.com/alphabetical/?per_page=100"

urls_all_pages <- c(str_c(base_url,
                          "&page=",
                          1:21)
                     )

all_pages <- map(urls_all_pages, scrape_page)

all_cheeses <- bind_rows(all_pages)

head(all_cheeses)
```

<!-- Grad level ideas: -->
<!-- - scrape something that isn't a table -->
