---
title: "`dplyr` Review"
editor: source
execute: 
  echo: true
  eval: false
  warning: false
  message: false
---

```{r}
#| label: set-up
#| include: false

library(tidyverse)
```

::: {.callout-caution}
# Optional Content

This module consists of readings reviewing material typically taught in STAT
331. It is possible you can skip over portions of this reading. It is your
responsibility to decide which areas you need to review before diving into
Stat 541.

Answer the following questions to see if you can safely skip this section. 
:::

1. In essence, a `data.frame` is simply a special `list` - with a few extra 
restrictions on the `list` format.

Think about the datasets you have already worked with. Which of the following
restrictions on a `list` do you think are needed for the `list` to be a
`data.frame`? (Select all that apply)

a)  The elements of the list must all be vectors of the same length.
b)  The elements of the list must all be the same data type.
c)  The elements of the list must all have no missing values.
d)  The elements of the list must all have names.

2. Tibbles are described as "opinionated" dataframes. Which of the following are
true about a `tibble`'s behavior? (Select all that apply)

a)  `tibble`s only print the first 10 rows of a dataset
b)  `tibble`s allow for non-syntactic variable names, like `:)`
c)  `tibble`s never convert strings to factors
d)  `tibble`s create row names

*If you had a hard time answering these questions, I would recommend reviewing
[@sec-data-structures].*

3. Match each of the base R code excerpt to the associated dplyr verb. 

::: columns
::: {.column width="40%"}
- `filter()`

- `select()`

- `mutate()`

:::

::: {.column width="10%"}
:::

::: {.column width="50%"}
- `arrange()`

- `summarize()`

- `group_by()`
:::
:::

a. `penguins[order(penguins$bill_length_mm) , ]`

b. `penguins[penguins$species == "Adelie", ]`

c. `aggregate(bill_length_mm ~ species, data = penguins, FUN = mean)`

d. `with(penguins, mass_ratio = body_mass_g / flipper_length_mm)`

e. `penguins$species`

f. `mean(penguins[penguins$species == "Adelie", ], na.rm = TRUE)`

*If you had a hard time answering this question, I would recommend reviewing
[@sec-dplyr].*

4. Suppose we would like to study how the ratio of bill length to bill depth
across the different penguin species. Arrange the following steps into an order
that accomplishes this goal (assuming the steps are connected with a `|>` or 
a `%>%`).

```{r}
#| include: false

penguins |>
  mutate(
    bill_ratio = bill_length_mm / bill_depth_mm
  ) |>
  group_by(species) |>
  summarize(
    avg_bill_ratio = mean(bill_ratio, na.rm = TRUE)
  ) |>
  arrange(avg_bill_ratio)
```

```{r}
#| eval: false
#| echo: true

# a
arrange(avg_bill_ratio)

# b
group_by(species)

# c
penguins 

# d
summarize(
    avg_bill_ratio = mean(bill_ratio, na.rm = TRUE)
    )
  
# e
mutate(
    bill_ratio = bill_length_mm / bill_depth_mm
    )
  
```

*If you had a hard time answering this question, I would recommend reviewing
[@sec-pipe].*

## `dplyr`

You should feel comfortable using:

-   The five main dplyr verbs:

    -   `filter()`

    -   `arrange()`

    -   `select()`

    -   `mutate()`

    -   `summarize()`

-   Incorportaing `group_by()` to perform groupwise operations

-   Chaining together data wrangling operations with the pipe operator (`|>` or
`%>%`)

### Data Structures {#sec-data-structures}

Choose one of these two options:

::: columns
::: {.column width="45%"}

📽 [Watch a Video: Data Types in R](https://youtu.be/vGY5i_J2c-c?si=ERuoY8qZOZ1uVbuL)
:::

::: {.column width="45%"}
📖 [Read a Textbook: Data Types in R ](https://bookdown.org/rdpeng/rprogdatascience/r-nuts-and-bolts.html#r-objects)
:::
:::

In addition, read the following section from the first edition of R for DS:

📖 [Recommended Reading: Tibbles](https://r4ds.had.co.nz/tibbles.html)

### Data Wrangling with dplyr {#sec-dplyr}

If you had a hard time answering Question 3, I would recommend reviewing
this content.

📖 [Recommended Reading: Data Transformation](https://r4ds.hadley.nz/data-transform.html)

## 

📽 [`filter()` -- 11 minutes](https://youtu.be/pPI-prhLYqI)

📽 [`select()` -- 6 minutes](https://youtu.be/OInnmiccHt8)

📽 [`mutate()` -- 9 minutes](https://youtu.be/QWyUnQ2EzPs)

📽 [`arrange()` -- 7 minutes](https://youtu.be/4Qhdc81mpvo)

📽 [`summarize()`-- 8 minutes](http://somup.com/crflYibWZw)

📽 [`group_by()` -- 6 minutes](http://somup.com/crflqbbWFy)

### The Pipe Operator {#sec-pipe}

If you had a hard time answering Question 4, I would recommend also reviewing
this content.

📽 [Recommended Video: The pipe operator -- 8 minutes](http://somup.com/crflYfbWZD)

<!-- 📽 [Recommended Video: Extending the pipe -- 3 minutes](http://somup.com/crflYmbWup) -->
