---
title: "Quarto Dashboards, Part 1"
editor: source
execute: 
  eval: false
---

This week, we're taking the visualization skills we learned last week and using
them to create an informational dashboard. Quarto dashboards were integrated
into version 1.4, so you can use a familiar tool to create these dashboards. 
In Part 1, we will focus on creating "static" dashboards---where the user 
**does not** interact with the content. 

## Step 1 - What is a Quarto dashboard?

::: {.callout-required-video}
::: youtube-video-container

{{< video https://www.youtube.com/embed/HW7QbqI4fH0?si=Rw6Cwy173CNJN3M >}}
:::

</br>

- Slides for this Video: <https://mine.quarto.pub/quarto-dashboards/1-hello-dashboards/>
- Starter Documents: <https://github.com/mine-cetinkaya-rundel/olympicdash/blob/main/olympicdash-r-1.qmd>
:::

::: {.callout-check-in}
1. To make a Quarto dashboard what do you put in the `format` line of your YAML?

2. If you use the default options for a Quarto dashboard and your code makes 
two plots, how will they be output?

- Stacked horizontally (side-by-side)
- Stacked vertically

3. How would you define a card title in an R code chunk?
:::


## Step 2 - How do you make a Quarto dashboard?

::: {.callout-required-video}
::: youtube-video-container

{{< video https://www.youtube.com/embed/KdsQgwaY950?si=dS2dYH_hmrpeDjSC >}}
:::

</br>

- Slides for this Video: <https://mine.quarto.pub/quarto-dashboards/2-dashboard-components/>
- Starter Documents: <https://github.com/mine-cetinkaya-rundel/olympicdash/blob/main/olympicdash-r-2.qmd>
:::

::: {.callout-check-in}
4. In a Quarto dashboard what does a `#` create?

- A new column
- A new row
- A new page

Suppose your Quarto dashboard has a YAML that looks something like this:

<!-- ``` {.markdown filename="my-first-dash.qmd" style="overflow-y: hidden"} -->

```{.yaml}
---
title: "Olympic Games"
format: 
  dashboard:
    orientation: columns
---
```

5. With this YAML, what does a `##` create?

- A new column
- A new row
- A new page

6. What does a `###` create? 

- A new column
- A new row
- A new page

7. How do you change the size of a row? Of a column?

<!-- row: {height="70%"} -->
<!-- column: {width="70%"} -->

8. Can different pages of a Quarto dashboard use different orientations?
Meaning, can one page use a row orientation and another page use a column 
orientation?

9. Which of the following cells will become a card in a dashboard?

::: columns
::: {.column width="48%"}

a. 

```{r}
#| echo: fenced
x <- 2
```

b. 

```{r}
#| echo: fenced
plot(cars)
```

:::

::: {.column width="2%"}
:::

::: {.column width="48%"}
c. 

```{r}
#| echo: fenced
#| output: false

2 + 2
```

d. 

```{r}
#| echo: fenced
library(palmerpenguins)
```

:::
:::

10. How do you create a card outside of an R script? Meaning, how would you 
create a card in Quarto?

<!-- ::: {.card} -->
:::

# Additional Resources

::: {.callout-learn-more} 

- [Quarto Dashboard Gallery](https://quarto.org/docs/gallery/#dashboards) - Example dashboards from the community of Quarto users
- [Quarto Dashboard Tutorial](https://quarto.org/docs/dashboards/) - Simple starting guide
- [Quarto Dashboard Reference](https://quarto.org/docs/reference/formats/dashboard.html) - Reference to all functions / aspects of a Quarto dashboards
:::
