---
title: "Functions, Iteration & Efficiency"
execute: 
  eval: false
---

::: {.callout-important}
# Check-in: Iteration

6. Write a function called `summ_stats()` that takes a numeric vector `x` as
input and returns the mean, median, standard deviation, and IQR as a data frame.
You can use `tibble()` to create the data frame.
    * Example: `tibble(a = 1:2, b = 2:3)` creates a data frame with variables
    `a` and `b`.


```{r}
#| label: ex_summ_stats_soln1
#| echo: false

summ_stats <- function(x) {
    tibble(
        mean = mean(x, na.rm = TRUE),
        median = median(x, na.rm = TRUE),
        sd = sd(x, na.rm = TRUE),
        iqr = IQR(x, na.rm = TRUE)
    ) 
}
```

7. Use `map()` to apply your `summ_stats()` function to the numeric columns in 
the `diamonds` dataset. *Hint:* Look up the `bind_rows()` documentation from
dplyr to combine summary statistics for all quantitative variables into one
data frame. The `.id` argument will be especially helpful in adding the variable
names!

```{r}
#| label: ex_summ_stats_soln2
#| echo: false
#| eval: false

diamonds %>% 
  select(where(is.numeric)) |> 
  map(.f = summ_stats) %>% 
  bind_rows(.id = "variable") 

```

8. Let's make the output more intuitive, with the variables on the columns and
the summary statistics on the rows. You will need to do a double pivot 
(`pivot_longer()` then `pivot_wider()`) to achieve this result!

```{r}
#| label: double-pivot
#| echo: false
#| eval: false

diamonds %>% 
  select(where(is.numeric)) %>% 
  bind_rows(.id = "variable") %>% 
  pivot_longer(cols = -variable, 
               names_to = "statistic", 
               values_to = "value") %>% 
  pivot_wider(names_from = variable, values_from = value)

```
:::


The function below uses `if()`, `else if()` and `else` to check how a number should be classified.

```{r}
#| echo: true
#| label: non-vectorized-function

pos_neg_zero <- function(x){
  if(x > 0){
    return("Greater than 0!")
  } 
  else if (x < 0){
    return("Less than 0!")
  } 
  else {
    return("Equal to 0!")
  }
}
```

These functions (`if()`, `else if()`) are not vectorized, meaning they cannot accept a vector as an input, which is why the code below returns an error:

```{r}
#| error: true
#| label: error-with-vector-for-non-vectorized-function
#| echo: true

x <- seq(from = -4, to = 6)
pos_neg_zero(x)
```

Since this function is not vectorized, we would need to use `map_chr()` to apply the `pos_neg_zero()` function to every element of the vector `x`:

```{r}
#| label: iterating-non-vectorized-function
#| echo: true

map_chr(x, .f = pos_neg_zero)
```

1.  Rewrite the `pos_neg_zero()` function so it is vectorized. Meaning, the function should be capable of accepting a vector as an input. *Hint: The trick **is not** adding `map_chr()` into the function body!*


<!-- need to have a question using map2 -->
